### 2019-01-15 submission/manuscript.Rmd
Fresh make write.paper

# Assessing the Differences in 16S rRNA Gene Sequencing Due to High Fidelity DNA Polymerase

# Influence of DNA Polymerase on PCR bias, chimera formation, and errors when sequencing 16S rRNA genes from complex communities

\begin{center}
\vspace{25mm}

@@

Co-author e-mails:

* marcsze@med.umich.edu

@@

@@

## Figure Lengends

@@

@@

@@

@@

@@

@@

@@

>>>>

\vspace{25mm}
\begin{center}

\textbf{\huge Assessing the differences in 16S rRNA gene sequence data that are due to choice of DNA polymerase and number of rounds of PCR}

\vspace{50mm}

\textbf{Running title:} Quantifying the choice of DNA polymerase

\vspace{25mm}

@@

@@

PCR amplification of 16S rRNA genes is a critical, yet under appreciated step in the generation of sequence data to describe the taxonomic composition of microbial communities. Numerous factors in the design of PCR can impact the sequencing error rate, the abundance of chimeric sequences, and the degree to which the fragments in the product represent their abundance in the original sample. The choice of PCR cycling conditions are known to impact the outcome of PCR and downstream inferences. We compared the performance of high fidelity polymerases and varying number of rounds of amplification when amplifying a mock community with known sequences and community structure and human stool samples. Although it is impossible to derive specific recommendations, we did observe general trends. Namely, using a polymerase with the highest possible fidelity and minimizing the number of rounds of PCR reduced the sequencing error rate, fraction of chimeric sequences, and bias. Evidence of bias at the sequence level was subtle and could not be ascribed to the fragments' fraction of bases that were guanines or cytosines. When analyzing the mock community data, the amount that the community deviated from the expected composition increased with rounds of PCR. The effect was inconsistent for human stool samples. Overall the results underscore the difficulty of comparing sequence data that are generated by different PCR protocols; however, the results indicate that the variation in human stool samples is generally larger than that introduced by the choice of polymerase or number of rounds of PCR.

### Importance

The growth in interest in characterizing microbial communities from diverse communities has exploded with the steep decline in sequencing costs. A significant amount of effort has gone into understanding the error profiles of DNA sequencers; however, little work has been done to understand the effects of choices that researchers make to generate the PCR amplicons that are sequenced. Here we quantified the effects of the choice of polymerase and number of PCR cycles on the quality of downstream data. Overall, we found that these choices can have a profound impact on the way that a microbial community is represented in the sequence data. Although the effects are relatively small compared to the variation in human stool samples, care should be taken to use polymerases with the highest possible fidelity and to minimize the number of rounds of PCR.

@@

@@

\includegraphics[width=0.9\columnwidth]{../results/figures/mock_error.pdf}

@@

\includegraphics[width=0.9\columnwidth]{../results/figures/chimera_plots.pdf}

@@

\includegraphics[width=0.9\columnwidth]{../results/figures/species_bias.pdf}

@@

\includegraphics[width=0.8\columnwidth]{../results/figures/mock_community.pdf}

@@

\includegraphics[width=0.9\columnwidth]{../results/figures/stool_community.pdf}

@@

\includegraphics[width=0.9\columnwidth]{../results/figures/drift.pdf}

@@

\includegraphics[width=0.9\columnwidth]{../results/figures/salmonella_bias.pdf}
